Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-24 manuscript/manuscript.tex
section 3

We can extend model (\ref{eq:msndens}) to the regression setting by modeling $\boldsymbol\zeta_{ki}$ as a function of covariates. Here we adopt a convenient stochastic representation of the MSN density (Azzalini and Dalla Valle, 1996):
\begin{equation}
\mathbf{y}_{i}|(z_i=k,t_i) = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k + \boldsymbol\epsilon_{ki}, \label{eq:msnreg}
\end{equation}
where 
$\mathbf{X}_i$ is a $J \times Jp$ design matrix that includes potential time-varying covariates (e.g., indicators denoting quarterly visits); $\boldsymbol{\beta}_k=(\beta_{k11},\ldots,\beta_{k1p},\ldots,\beta_{kJ1},\ldots,\beta_{kJp})^T$ is a $Jp\times 1$ vector of cluster- and outcome-specific regression coefficients; $t_i\sim N_{[0,\infty)}(0,1)$ is a subject-specific standard normal random variable truncated below by zero; $\boldsymbol\psi_k=(\psi_{k1},\ldots,\psi_{kJ})^T$ is a $J \times 1$ vector of cluster-specific skewness parameters; and $\boldsymbol{\epsilon}_{ki} \sim N_J(\boldsymbol0,\boldsymbol\Sigma_k)$ is a $J\times 1$ vector of error terms. Thus, conditional on $t_i$ and $z_i=k$, $\boldsymbol{y}_i$ is distributed as $N_J(\mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k, \boldsymbol{\Sigma}_k)$. Marginally (after integrating over $t_i$), $\mathbf{y}_i$ is distributed $MSN_J(\boldsymbol\zeta_{ki}, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$, where through back-transformation 
\begin{eqnarray*}
\boldsymbol\zeta_{ki} &=& \mathbf{X}_i\boldsymbol\beta_k\\
\boldsymbol\alpha_k &=& \frac{1}{\sqrt{1 - \boldsymbol\psi_k^T 
\boldsymbol\Omega^{-1}_k\boldsymbol\psi_k}} \boldsymbol\omega_k \boldsymbol\Omega^{-1}_k\boldsymbol\psi_k~~\text{and}\\
\boldsymbol\Omega_k &=& \boldsymbol\Sigma_k  + \boldsymbol\psi_k \boldsymbol\psi_k^T,
\end{eqnarray*}
where $\boldsymbol\omega_k = \text{Diag}(\sqrt{\omega_{k,11}},...,\sqrt{\omega_{k,JJ}})$ is the $J \times J$ diagonal matrix containing the square root of the diagonal entries of $\boldsymbol\Omega_k$. Additional details can be found in Fr\"{u}wirth-Schnatter and Pyne (2010).

@@

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,Y^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& N_{J-q}(\boldsymbol\mu^{miss}_{ki},\boldsymbol\Sigma^{miss}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{miss}_{ki}&=&\\
\boldsymbol\Sigma^{miss}_k&=& \nonumber
\end{eqnarray}
\textbf{Carter -- work on the above -- you will need to define notation and refer readers back to equation (3) as needed. We can discuss next week if needed}
These results follow from conventional multivariate normal theory. An attractive feature of this imputation algorithm is that it provides more opportunities to explore the parameter space than multiple imputation \textbf{[based on summary stats right?]} and avoids multiplicative run-time scaling in $m$, the number of imputations \textbf{Give refs}. In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over complete case analysis.

@@

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$.} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. \textbf{[See my Bayesian Analysis paper for guidance on this part]}.

\textbf{[I stopped here]}

$$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{k' \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}}$$

where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$

where ${c}_{ik} = \log \sum_{k' \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{k' \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $e^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{k' \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}} = \log \left ( 1 + \sum_{k' \notin \{k,K \}} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}} \right )$$

We can use the quantities to re-express the full conditionals for $\boldsymbol\delta_k$ as
$$p(\boldsymbol\delta_k|\mathbf{Z},\boldsymbol\delta_{k' \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}} \right )^{U_{ik}} \left (\frac{1}{1 + e^{\eta_{ik}}} \right )^{1-U_{ik}} = p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(e^{\eta_{ik}})^{U_{ik}}}{1 + e^{\eta_{ik}}}$$

>>>>

We can extend model (\ref{eq:msndens}) to the regression setting by modeling $\boldsymbol\zeta_{ki}$ as a function of covariates. Here we adopt a convenient stochastic representation of the MSN density (Azzalini and Dalla Valle, 1996):
\begin{equation}
\mathbf{y}_{i}|(z_i=k,t_i) = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k + \boldsymbol\epsilon_{ki}, \label{eq:msnreg}
\end{equation}
where 
$\mathbf{X}_i$ is a $J \times Jp$ design matrix that includes potential time-varying covariates (e.g., indicators denoting quarterly visits); $\boldsymbol{\beta}_k=(\beta_{k11},\ldots,\beta_{k1p},\ldots,\beta_{kJ1},\ldots,\beta_{kJp})^T$ is a $Jp\times 1$ vector of cluster- and outcome-specific regression coefficients; $t_i\sim N_{[0,\infty)}(0,1)$ is a subject-specific standard normal random variable truncated below by zero; $\boldsymbol\psi_k=(\psi_{k1},\ldots,\psi_{kJ})^T$ is a $J \times 1$ vector of cluster-specific skewness parameters; and $\boldsymbol{\epsilon}_{ki} \sim N_J(\boldsymbol0,\boldsymbol\Sigma_k)$ is a $J\times 1$ vector of error terms. Thus, conditional on $t_i$ and $z_i=k$, $\boldsymbol{y}_i$ is distributed as $N_J(\mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k, \boldsymbol{\Sigma}_k)$. Marginally (after integrating over $t_i$), $\mathbf{y}_i$ is distributed $MSN_J(\boldsymbol\zeta_{ki}, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$, where through back-transformation 
\begin{eqnarray*}
\boldsymbol\zeta_{ki} &=& \mathbf{X}_i\boldsymbol\beta_k,\\
\boldsymbol\alpha_k &=& \frac{1}{\sqrt{1 - \boldsymbol\psi_k^T 
\boldsymbol\Omega^{-1}_k\boldsymbol\psi_k}} \boldsymbol\omega_k \boldsymbol\Omega^{-1}_k\boldsymbol\psi_k,~~\text{and}\\
\boldsymbol\Omega_k &=& \boldsymbol\Sigma_k  + \boldsymbol\psi_k \boldsymbol\psi_k^T,
\end{eqnarray*}
where $\boldsymbol\omega_k = \text{Diag}(\sqrt{\omega_{k,11}},...,\sqrt{\omega_{k,JJ}})$ is the $J \times J$ diagonal matrix containing the square root of the diagonal entries of $\boldsymbol\Omega_k$. Additional details can be found in Fr\"{u}wirth-Schnatter and Pyne (2010).

@@

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,Y^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& N_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs})\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},~\text{where}\nonumber
\end{eqnarray}
$\boldsymbol\Sigma_k$ is partitioned into four sub-matrices $\boldsymbol\Sigma_{11},\boldsymbol\Sigma_{12},\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ such that $\boldsymbol\Sigma_{11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ are defined in the same manner. These results follow from conventional multivariate normal theory. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations. In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over complete case analysis.

@@

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$.} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. \textbf{[See my Bayesian Analysis paper for guidance on this part]}.
$$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}}$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $e^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )$$

We can use the quantities to re-express the full conditionals for $\boldsymbol\delta_k$ as
$$p(\boldsymbol\delta_k|\mathbf{Z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}} \right )^{U_{ik}} \left (\frac{1}{1 + e^{\eta_{ik}}} \right )^{1-U_{ik}} = p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(e^{\eta_{ik}})^{U_{ik}}}{1 + e^{\eta_{ik}}}$$
